---
title: "Univariate Time Series for NHR"
author: "Iliana Kleovoulou"
date: "2023-05-26"
output: html_document
---

```{r, echo=FALSE}
library(INLA)
library(brinla)
library(dplyr)
source("functions.R")
library(hrbrthemes)
library(gridExtra)
```

Import data:
```{r}
data = read.csv("Sensor_data3.csv", sep = ",")
data$Alert = as.factor(data$Alert)
data$Participant.Id = as.factor(data$Participant.Id)
```

```{r}
#subset data for patient 4.1
data4 = data[data$Participant.Id == "4.1", ]
n = nrow(data4)
# create train and test set
n.hold = 5
n.train = n - n.hold
train.df = data4[1:n.train, ]
test.df = tail(data4, n.hold)


pt4.x = train.df$time_new # with negative values 
pt4.x = c(pt4.x, rep(NA, n.hold)) # add NAs

pt4.x2 = 1:n

pt4.y = train.df$NightHeartRate
pt4.y = c(pt4.y, rep(NA, n.hold))

data.ar1.pt4 = cbind.data.frame(pt4.y, pt4.x, pt4.x2)

modelformula = pt4.y ~ 1 + pt4.x +  f(pt4.x2, model = "ar1")
p4.ar1.NHR = inla(formula = modelformula,
                    data = data.ar1.pt4,
                    family =  "gaussian",
                    control.predictor = list(compute = TRUE),
                    control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE),
                    control.family = list(initial = 10, fixed=TRUE)
                    )
summary(p4.ar1.NHR)
```


```{r}
sigma2.w.dist = inla.tmarginal(
  fun = function(x)
    exp(-x),
  marginal = p4.ar1.NHR$internal.marginals.hyperpar$`Log precision for pt4.x2`
)
sigma2.w.zm = inla.zmarginal(sigma2.w.dist)
```

```{r}
model.selection.criteria(p4.ar1.NHR)
```
````{r}
p4.ar1.NHR$mlik
````

```{r}
#CPO and PIT
cpoi = p4.ar1.NHR$cpo$cpo[!is.na(p4.ar1.NHR$cpo$cpo)]
cpo = -sum(log(cpoi))
piti = p4.ar1.NHR$cpo$pit[!is.na(p4.ar1.NHR$cpo$pit)]
pit=ks.test(piti,"punif",0,1)$statistic

print(paste0("CPO: ", round(cpo,4)))
print(paste0("PIT: ", round(pit,5)))
```


Check for autocorrelation in residuals:
```{r}
fit.ar1.p4.NHR = p4.ar1.NHR$summary.fitted.values$mean
resid.ar1.pt4.NHR = pt4.y - fit.ar1.p4.NHR # calculate residuals

acf.ar1.pt4.NHR = acf(resid.ar1.pt4.NHR, plot = T,na.action = na.pass)
```

Now we fit for another patient (patient 5) which belongs to the non alert group

```{r}
#subset data for patient 4.1
data5 = data[data$Participant.Id == "5.1", ]
data5 = data5[data5$time_new < 0,]
n = nrow(data5)
# create train and test set
n.hold = 5
n.train = n - n.hold
train.df = data5[1:n.train, ]
test.df = tail(data5, n.hold)


pt5.x = train.df$time_new # with negative values 
pt5.x = c(pt5.x, rep(NA, n.hold)) # add NAs

pt5.x2 = 1:n

pt5.y = train.df$NightHeartRate
pt5.y = c(pt5.y, rep(NA, n.hold))

data.ar1.pt5 = cbind.data.frame(pt5.y, pt5.x, pt5.x2)

modelformula = pt5.y ~ 1 + pt5.x +  f(pt5.x2, model = "ar1")
p5.ar1.NHR = inla(formula = modelformula,
                    data = data.ar1.pt5,
                    family =  "gaussian",
                    control.predictor = list(compute = TRUE),
                    control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE),
                    control.family = list(initial = 10, fixed=TRUE)
                    )
summary(p5.ar1.NHR)
```

```{r}
model.selection.criteria(p5.ar1.NHR)
```
```{r, echo=FALSE}
#CPO and PIT
cpoi = p5.ar1.NHR$cpo$cpo[!is.na(p5.ar1.NHR$cpo$cpo)]
cpo = -sum(log(cpoi))
piti = p5.ar1.NHR$cpo$pit[!is.na(p5.ar1.NHR$cpo$pit)]
pit=ks.test(piti,"punif",0,1)$statistic

print(paste0("CPO: ", round(cpo,4)))
print(paste0("PIT: ", round(pit,5)))
```

Check for autocorrelation in residuals:
```{r}
fit.ar1.p5.NHR = p5.ar1.NHR$summary.fitted.values$mean
resid.ar1.pt5.NHR = pt5.y - fit.ar1.p5.NHR # calculate residuals

acf.ar1.pt5.NHR = acf(resid.ar1.pt5.NHR, plot = T,na.action = na.pass)
```
Fit this model to all patients (30 separated models) and extract their coefficients.

```{r, echo=FALSE}
ids = data$Participant.Id
df = data.frame()
state = list()
rho = list()

for (i in unique(ids)){
  data.i = data[data$Participant.Id == i, ]
  data.i = data.i[data.i$time_new < 0,] # only the negative values
  
  n = nrow(data.i) # number of observations 
  
  n.hold = 5
  n.train = n - n.hold
  train.df = data.i[1:n.train, ]
  test.df = tail(data.i, n.hold)


  x = train.df$time_new # with negative values 
  x = c(x, rep(NA, n.hold)) # add NAs
  x2 = 1:n

  y = train.df$NightHeartRate
  y = c(y, rep(NA, n.hold))

  data.ar1.i = cbind.data.frame(y, x, x2)
  
  modelformula = y ~ 1 + x +  f(x2, model = "ar1")
  modelfamily = "gaussian"
  ar1 = inla(formula = modelformula,
                      data = data.ar1.i,
                      family =  modelfamily,
                      control.predictor = list(compute = TRUE),
                      control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE),
                      control.family = list(initial = 10, fixed=TRUE)
                      )
  
  data.ar1.i$x = as.numeric(data.ar1.i$x)
  fit = ar1$summary.fitted.values
  
  sigma2.w.dist = inla.tmarginal(
  fun = function(x)
    exp(-x),
  marginal = ar1$internal.marginals.hyperpar$`Log precision for x2`)
  output.sigma2 = inla.zmarginal(sigma2.w.dist)$mean
 
  
  output.int = ar1$summary.fixed$mean[1]
  output.trend = ar1$summary.fixed$mean[2]
  output.coef = ar1$summary.hyperpar$mean[2]
  
  state = append(state,list(ar1$internal.marginals.hyperpar$`Log precision for x2`))
  rho = append(rho, list(ar1$internal.marginals.hyperpar$`Rho_intern for x2`))
  
  output = c(output.int, output.trend, output.coef, output.sigma2)
  
  df = rbind(df, output)
}
colnames(df) = c("Level", "Trend", "AR coefficient", "StateVariance")
df$ID = unique(ids)
```

```{r, echo=FALSE}
# patients ids with alert
alert.ids = c("4.1", "11.1", "12.1", "30.1", "51.1", "58.1", "68.1", 
              "69.1", "89.1", "94.1", "95.1", "127.1")

non.alerts.ids = c("5.1", "19.1", "20.6", "26.1", "32.1", "33.1", "35.1", "39.1",
                   "46.1", "53.1", "54.1", "72.1", "76.1", "79.1", "84.1", "92.1",
                   "109.1", "113.1", "20.1")
df.alert = df[df$ID %in% alert.ids, ]
df.nonalert = df[df$ID %in% non.alerts.ids,  ]
```

```{r, echo=FALSE}
df$alert = ifelse(df$ID %in% alert.ids, 1, 0)
df$alert = as.factor(df$alert)
```

```{r}
coefficients_long_nhr <- tidyr::pivot_longer(df[,1:4], cols = everything(), names_to = "Coefficient", values_to = "Value")


# Calculate mean and median for each coefficient
summary_stats <- coefficients_long_nhr %>%
  group_by(Coefficient) %>%
  summarise(Mean = mean(Value),
            Median = median(Value),
            CI_lower = quantile(Value, 0.025),
            CI_upper = quantile(Value, 0.975))
 
summary_stats
```

Now we can fit Autoregressive of order 2:
```{r}
# for patient 4
modelformula.ar2 = pt4.y ~ 1 + pt4.x +  f(pt4.x2, model = "ar", order = 2)
                                          
patient4.ar2 = inla(formula = modelformula.ar2,
                    data = data.ar1.pt4,
                    family =  "gaussian",
                    control.predictor = list(compute = TRUE),
                    control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE),
                    control.family = list(initial = 10, fixed=TRUE)
                    )
summary(patient4.ar2)
```
```{r}
pacf = patient4.ar2$summary.hyperpar$mean[2:3]
phi = inla.ar.pacf2phi(pacf)
print(phi)
```
```{r}
pacf = patient4.ar2$summary.hyperpar$'0.025quant'[2:3]
phi = inla.ar.pacf2phi(pacf)
print(phi)
```
```{r}
pacf = patient4.ar2$summary.hyperpar$'0.975quant'[2:3]
phi = inla.ar.pacf2phi(pacf)
print(phi)
```



```{r, echo=FALSE}
psiwt.ar2 <- ARMAtoMA(ar = phi, ma = 0, 100)
precision.x.hat <- patient4.ar2$summary.hyperpar$mean[1]

sigma2.w.hat <-
  inla.emarginal(
    fun = function(x)
      exp(-x),
    marginal = patient4.ar2$internal.marginals.hyperpar$`Log precision for pt4.x2`
  ) / sum(psiwt.ar2 ^ 2)

cat(paste(
  "Estimated state noise variance, sigma2.w",
  round(sigma2.w.hat, 3),
  sep = " = "
),
"\n")
```
```{r}
sigma2.w.hat.t <-
  inla.tmarginal(
    fun = function(x)
      exp(-x),
    marginal = patient4.ar2$internal.marginals.hyperpar$`Log precision for pt4.x2`
  ) / sum(psiwt.ar2 ^ 2)

a = inla.zmarginal(sigma2.w.hat.t)
```

```{r}
model.selection.criteria(patient4.ar2)
```


```{r, echo=FALSE}
#CPO and PIT
cpoi = patient4.ar2$cpo$cpo[!is.na(patient4.ar2$cpo$cpo)]
cpo = -sum(log(cpoi))
piti = patient4.ar2$cpo$pit[!is.na(patient4.ar2$cpo$pit)]
pit=ks.test(piti,"punif",0,1)$statistic

print(paste0("CPO: ", round(cpo,4)))
print(paste0("PIT: ", round(pit,5)))
```
Check residuals:
```{r, echo=FALSE}
fit.ar2trend.pt4 = patient4.ar2$summary.fitted.values$mean
resid.ar2trend.pt4 = pt4.y - fit.ar2trend.pt4
acf.ar2trend.pt4 = acf(resid.ar2trend.pt4, plot = T, na.action = na.pass)
```
Fit for alert free patients:
```{r}
modelformula = pt5.y ~ 1 + pt5.x +  f(pt5.x2, model = "ar", order = 2)
                                          
modelfamily = "gaussian"
patient5.ar2 = inla(formula = modelformula,
                    data = data.ar1.pt5,
                    family =  modelfamily,
                    control.predictor = list(compute = TRUE),
                    control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE),
                    control.family = list(initial = 10, fixed=TRUE)
                    )
summary(patient5.ar2)
```

```{r}
model.selection.criteria(patient5.ar2)
```
```{r, echo=FALSE}
#CPO and PIT
cpoi = patient5.ar2$cpo$cpo[!is.na(patient5.ar2$cpo$cpo)]
cpo = -sum(log(cpoi))
piti = patient5.ar2$cpo$pit[!is.na(patient5.ar2$cpo$pit)]
pit=ks.test(piti,"punif",0,1)$statistic

print(paste0("CPO: ", round(cpo,4)))
print(paste0("PIT: ", round(pit,5)))
```
Use inverse transformation to retrieve phis.
```{r}
# recover phis using the inverse transformation
pacf = patient5.ar2$summary.hyperpar$mean[2:3]
phi = inla.ar.pacf2phi(pacf)
print(phi)
```

```{r}
fit.ar2trend.pt5 = patient5.ar2$summary.fitted.values$mean
resid.ar2trend.pt5 = pt5.y - fit.ar2trend.pt5
acf.ar2trend = acf(resid.ar2trend.pt5, plot = T, na.action = na.pass)
```
Now we can fit for all patients:

```{r, echo=FALSE}
ids = data$Participant.Id
df = data.frame()
state = list()
rho1 = list()
rho2 = list()

for (i in unique(ids)){
  data.i = data[data$Participant.Id == i, ]
  data.i = data.i[data.i$time_new < 0,] # only the negative values
  
  n = nrow(data.i) # number of observations 
  
  n.hold = 5
  n.train = n - n.hold
  train.df = data.i[1:n.train, ]
  test.df = tail(data.i, n.hold)


  x = train.df$time_new # with negative values 
  x = c(x, rep(NA, n.hold)) # add NAs
  x2 = 1:n

  y = train.df$NightHeartRate
  y = c(y, rep(NA, n.hold))

  data.ar2.i = cbind.data.frame(y, x, x2)
  
  modelformula = y ~ 1 + x +  f(x2, model = "ar", order = 2)
  modelfamily = "gaussian"
  ar2 = inla(formula = modelformula,
                      data = data.ar2.i,
                      family =  modelfamily,
                      control.predictor = list(compute = TRUE),
                      control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE),
                      control.family = list(initial = 10, fixed=TRUE)
                      )
  
  data.ar2.i$x = as.numeric(data.ar2.i$x)
  fit = ar2$summary.fitted.values
  
  sigma2.w.dist = inla.tmarginal(
  fun = function(x)
    exp(-x),
  marginal = ar2$internal.marginals.hyperpar$`Log precision for x2`)
  output.sigma2 = inla.zmarginal(sigma2.w.dist)$mean
 
  
  output.int = ar2$summary.fixed$mean[1]
  output.trend = ar2$summary.fixed$mean[2]
  output.coef1 = ar2$summary.hyperpar$mean[2]
  output.coef2 = ar2$summary.hyperpar$mean[3]
  
  
  state = append(state,list(ar2$internal.marginals.hyperpar$`Log precision for x2`))
  rho1 = append(rho1, list(ar2$internal.marginals.hyperpar$`Intern PACF1 for pt5.x2`))
  rho2 = append(rho2, list(ar2$internal.marginals.hyperpar$`Intern PACF2 for pt5.x2`))

  output = c(output.int, output.trend, output.coef1, output.coef2, output.sigma2)
  
  df = rbind(df, output)
}
colnames(df) = c("Level", "Trend", "Rho1", "Rho2", "StateVariance")
df$ID = unique(ids)
```

We need to transform phis:
```{r}
phis = inla.ar.pacf2phi(df[3:4])
phis = as.data.frame(phis)
colnames(phis) = c("Phi1", "Phi2")
df = cbind(df, phis)

```

Summary of the results
```{r}
coefficients_long_nhr_ar2 <- tidyr::pivot_longer(df[,c(1,2,5,7,8)], cols = everything(), names_to = "Coefficient", values_to = "Value")

# Calculate mean and median for each coefficient
summary_stats_ar2 <- coefficients_long_nhr_ar2 %>%
  group_by(Coefficient) %>%
  summarise(Mean = mean(Value),
            Median = median(Value),
            CI_lower = quantile(Value, 0.025),
            CI_upper = quantile(Value, 0.975))
 
summary_stats_ar2

```


Forecasting:
First we do forecasting for patient 4 for AR(1) model and AR(2) model
```{r, echo=FALSE}
n = nrow(data4)
n.hold = 5
n.train = n - n.hold

fore =c(rep(NA, n.train),
  tail(p4.ar1.NHR$summary.fitted.values$mean, n.hold))
obs = data4$NightHeartRate
fore.low <- c(
  rep(NA, n.train),
  tail(p4.ar1.NHR$summary.fitted.values$mean, n.hold) -
    2 * tail(p4.ar1.NHR$summary.fitted.values$sd, n.hold)
)
fore.up <- c(
  rep(NA, n.train),
  tail(p4.ar1.NHR$summary.fitted.values$mean, n.hold) +
    2 * tail(p4.ar1.NHR$summary.fitted.values$sd, n.hold)
)

plot.df <-
  cbind.data.frame(time = 1:n, obs, fore, fore.low, fore.up)

plot.ar1 = plot.df %>%
  ggplot(aes(x = time)) +
  geom_line(aes(y = obs), color = "#D55E00") +
  geom_line(aes(y = fore), color = "#0072B2") +
  geom_ribbon(aes(ymin = fore.low, ymax = fore.up), alpha =
                .2) +
  geom_vline(xintercept = n.train + 1,
             ylab = "",
             size = 0.2) +
  xlab(label = "t") +
  ylab(label = "Night Heart Rate") +
  ggtitle("AR(1) with level and trend model for pt4")

fore =c(rep(NA, n.train),
  tail(patient4.ar2$summary.fitted.values$mean, n.hold))
obs = data4$NightHeartRate
fore.low <- c(
  rep(NA, n.train),
  tail(patient4.ar2$summary.fitted.values$mean, n.hold) -
    2 * tail(patient4.ar2$summary.fitted.values$sd, n.hold)
)
fore.up <- c(
  rep(NA, n.train),
  tail(patient4.ar2$summary.fitted.values$mean, n.hold) +
    2 * tail(patient4.ar2$summary.fitted.values$sd, n.hold)
)

plot.df <-
  cbind.data.frame(time = 1:n, obs, fore, fore.low, fore.up)
plot.ar2 = plot.df %>%
  ggplot(aes(x = time)) +
  geom_line(aes(y = obs), color = "#D55E00") +
  geom_line(aes(y = fore), color = "#0072B2") +
  geom_ribbon(aes(ymin = fore.low, ymax = fore.up), alpha =
                .2) +
  geom_vline(xintercept = n.train + 1,
             ylab = "",
             size = 0.2) +
  xlab(label = "t") +
  ylab(label = "Night Heart Rate") +
  ggtitle("AR(2) with level and trend model for pt4")


grid.arrange(plot.ar1,plot.ar2)
```

MAE and MAPE
```{r}
mae.ar1 = mae(yhold = tail(data4$NightHeartRate,n.hold), 
              yfore = tail(p4.ar1.NHR$summary.fitted.values$mean, n.hold))
mape.ar1 = mape(yhold = tail(data4$NightHeartRate,n.hold), 
              yfore = tail(p4.ar1.NHR$summary.fitted.values$mean, n.hold))
print(paste("For AR(1) using data for pt.4 we have: ", mae.ar1, "and", mape.ar1))
```

```{r}
mae.ar2 = mae(yhold = tail(data4$NightHeartRate,n.hold), 
              yfore = tail(patient4.ar2$summary.fitted.values$mean, n.hold))
mape.ar2 = mape(yhold = tail(data4$NightHeartRate,n.hold), 
              yfore = tail(patient4.ar2$summary.fitted.values$mean, n.hold))
print(paste("For AR(2) using data for pt.4 we have: ", mae.ar2, "and", mape.ar2))
```

For patient 5:
```{r, echo=FALSE}
n = nrow(data5)
n.hold = 5
n.train = n - n.hold

fore =c(rep(NA, n.train),
  tail(p5.ar1.NHR$summary.fitted.values$mean, n.hold))
obs = data5$NightHeartRate
fore.low <- c(
  rep(NA, n.train),
  tail(p5.ar1.NHR$summary.fitted.values$mean, n.hold) -
    2 * tail(p5.ar1.NHR$summary.fitted.values$sd, n.hold)
)
fore.up <- c(
  rep(NA, n.train),
  tail(p5.ar1.NHR$summary.fitted.values$mean, n.hold) +
    2 * tail(p5.ar1.NHR$summary.fitted.values$sd, n.hold)
)

plot.df <-
  cbind.data.frame(time = 1:n, obs, fore, fore.low, fore.up)

plot.ar1 = plot.df %>%
  ggplot(aes(x = time)) +
  geom_line(aes(y = obs), color = "#D55E00") +
  geom_line(aes(y = fore), color = "#0072B2") +
  geom_ribbon(aes(ymin = fore.low, ymax = fore.up), alpha =
                .2) +
  geom_vline(xintercept = n.train + 1,
             ylab = "",
             size = 0.2) +
  xlab(label = "t") +
  ylab(label = "Night Heart Rate") +
  ggtitle("AR(1) with level and trend model for pt5")

fore =c(rep(NA, n.train),
  tail(patient5.ar2$summary.fitted.values$mean, n.hold))
obs = data5$NightHeartRate
fore.low <- c(
  rep(NA, n.train),
  tail(patient5.ar2$summary.fitted.values$mean, n.hold) -
    2 * tail(patient5.ar2$summary.fitted.values$sd, n.hold)
)
fore.up <- c(
  rep(NA, n.train),
  tail(patient5.ar2$summary.fitted.values$mean, n.hold) +
    2 * tail(patient5.ar2$summary.fitted.values$sd, n.hold)
)

plot.df <-
  cbind.data.frame(time = 1:n, obs, fore, fore.low, fore.up)
plot.ar2 = plot.df %>%
  ggplot(aes(x = time)) +
  geom_line(aes(y = obs), color = "#D55E00") +
  geom_line(aes(y = fore), color = "#0072B2") +
  geom_ribbon(aes(ymin = fore.low, ymax = fore.up), alpha =
                .2) +
  geom_vline(xintercept = n.train + 1,
             ylab = "",
             size = 0.2) +
  xlab(label = "t") +
  ylab(label = "Night Heart Rate") +
  ggtitle("AR(2) with level and trend model for pt5")

grid.arrange(plot.ar1,plot.ar2)
```

MAE and MAPE
```{r}
mae.ar1 = mae(yhold = tail(data5$NightHeartRate,n.hold), 
              yfore = tail(p5.ar1.NHR$summary.fitted.values$mean, n.hold))
mape.ar1 = mape(yhold = tail(data5$NightHeartRate,n.hold), 
              yfore = tail(p5.ar1.NHR$summary.fitted.values$mean, n.hold))
print(paste("For AR(1) using data for pt.5 we have: ", mae.ar1, "and", mape.ar1))
```

```{r}
mae.ar2 = mae(yhold = tail(data5$NightHeartRate,n.hold), 
              yfore = tail(patient5.ar2$summary.fitted.values$mean, n.hold))
mape.ar2 = mape(yhold = tail(data5$NightHeartRate,n.hold), 
              yfore = tail(patient5.ar2$summary.fitted.values$mean, n.hold))
print(paste("For AR(2) using data for pt.5 we have: ", mae.ar2, "and", mape.ar2))
```












