---
title: "VAR"
author: "Iliana Kleovoulou"
date: "2023-06-19"
output: html_document
---
# Gaussian Multivariate Time Series


```{r, echo=FALSE}
library("INLA")
source("functions.R")
library("dplyr")
library("stringr") 
library("lattice")
library("car")
library("mvtnorm")
```

```{r}
data = read.csv("Sensor_data3.csv", sep = ",")
data$Participant.Id = as.factor(data$Participant.Id)
data$Alert = as.factor(data$Alert)
```

In this section we model Gaussian Multivariate Time Series. 
We follow the section 12.1 of the time series book, and we fit
2 time series using independent AR(1) processes. To do that, we must first
create the indices which play an important role in the implementation of this model.
This model is using univariate time series, so we use the data from one patient to fit it.
```{r}
# subset data for patient 4.1
data.pt4 = data[data$Participant.Id == "4.1", ]
n = nrow(data.pt4)
n
# create train and test set
n.hold = 5
n.train = n - n.hold
train.df = data.pt4[1:n.train, ]
test.df = tail(data.pt4, n.hold)

N = 2*n
```

* First we need to create Y. Y is a vector contained the 2 dimensional time series,
stacked one below the other so Y is a stacked vector of time series.
* Then, we need to create two variables to denote the two intercepts, one for each sensor
* And 2 AR(1) structures to account for the correlation of each sensor.
* After that, we bind these columns together to create the data frame that we use for the model.
```{r}
# S3/S1 ratio
S3S1.pt4 = train.df$S3overS1*100
# Night Heart Rate
NHR.pt4 = train.df$NightHeartRate

# create the Y vector
Y = c(c(S3S1.pt4, rep(NA, n.hold)), c(NHR.pt4, rep(NA, n.hold)))

# intercepts
alpha.S3S1 = c(rep(1,n), rep(NA,n))
alpha.NHR = c(rep(NA,n), rep(1,n))

# trends
time.S3S1 <- c(data.pt4$time_new, rep(NA, n))
time.NHR <- c(rep(NA, n), data.pt4$time_new)

# AR(1) processes
b0.S3S1 = c(1:n, rep(NA, n))
b0.NHR = c(rep(NA, n), 1:n)

# for covariance matrix
t.index = 1:N
pt4.mvn.data = cbind.data.frame(Y,
                     alpha.S3S1, alpha.NHR,
                     b0.S3S1, b0.NHR,
                     time.S3S1, time.NHR,
                     t.index)

```

Now we are able to fit the model:
```{r}
formula.mvn.pt4 = Y ~ -1 + alpha.S3S1 + alpha.NHR + time.S3S1 + time.NHR + 
  f(t.index, model="iid2d", n=N) + 
  f(b0.S3S1, model = "ar1") +
  f(b0.NHR, model = "ar1")

inla.mvn.pt4 = inla(formula = formula.mvn.pt4,
                    family = c("gaussian"),
                    data = pt4.mvn.data,
                    control.inla = list(h = 1e-5, tolerance = 1e-3),
                    control.compute = list(dic = TRUE,waic = TRUE, cpo = TRUE,config = TRUE),
                    control.predictor = list(compute = TRUE),
                    control.family =list(hyper = list(prec = list(initial = 10,fixed = TRUE))))
result = inla.rerun(inla.mvn.pt4)
summary(result)
```
```{r, echo=FALSE}
library(brinla)
bri.hyperpar.summary(result)
```
```{r}
sigma.dist = inla.tmarginal(
  fun = function(x)
    exp(-x/2),
  marginal = result$internal.marginals.hyperpar$`Log precision for b0.NHR`
)
sigma.zm = inla.zmarginal(sigma.dist)
```
Posteriors of hyper-parameters
```{r, echo=FALSE}
bri.hyperpar.plot(result, together = F)
```
```{r}
model.sel.biv = model.selection.criteria(result)
model.sel.biv
```
```{r, echo=FALSE}
#CPO and PIT
cpoi = result$cpo$cpo[!is.na(result$cpo$cpo)]
cpo = -sum(log(cpoi))
piti = result$cpo$pit[!is.na(result$cpo$pit)]
pit=ks.test(piti,"punif",0,1)$statistic

print(paste0("CPO: ", round(cpo,4)))
print(paste0("PIT: ", round(pit,5)))
```
# Fitted values and forecast
To extract the fitted values from the multivariate model, we can't just take the
fitted values like we did before in the univariate case. Now we need to sample from
the predictive distributions of the responses using the function inla.posterior.sample().
This function provides M posterior samples for the model parameters and hyperparameters. 
After we get those, we can simulate fitted responses for each 1,...,M samples.For each of the  
2 sensors, the sample average of these  M fitted response values for each time  t
constitutes a fitted time series, which can then be plotted against the observed values.

```{r}
post.sample = inla.posterior.sample(1000, result)
```

Here is a function to calculate the fitted values
```{r}
fit.post.sample = function(x) {
  # firstly, we construct matrix V: V is a 2x2 covariance matrix 
  sigma2.v1 = 1 / x$hyperpar[1] # marginal variance for u_1
  sigma2.v2 = 1 / x$hyperpar[2] # marginal variance for u_2
  rho.v1v2 = x$hyperpar[3] # correlation
  cov.v1v2 = rho.v1v2 * sqrt(sigma2.v1) * sqrt(sigma2.v2) # covariance
  sigma.v =
    matrix(c(sigma2.v1, cov.v1v2, cov.v1v2, sigma2.v2), nrow = 2) # the covariance matrix
  
  # now we sample from the Multivariate Normal with mean 0 and covariance matrix the V
  V = rmvnorm(n, mean = c(0, 0), sigma = sigma.v)
  V1 = V[, 1] # first column for sensor 1 (S3/S1)
  V2 = V[, 2] # second column for sensor 2 (NHR)
  
  # using the M posterior samples, we simulate fitted responses for the 2 time series
  # For S3/S1
  fit.S3S1 = x$latent[grep("alpha.S3S1", rownames(x$latent))] +
    x$latent[grep("b0.S3S1", rownames(x$latent))] +
    x$latent[grep("time.S3S1", rownames(x$latent))] * 
    data.pt4$time_new +
    V1
  # for NHR
  fit.NHR = x$latent[grep("alpha.NHR", rownames(x$latent))] +
    x$latent[grep("b0.NHR", rownames(x$latent))] +
    x$latent[grep("time.NHR", rownames(x$latent))] * 
    data.pt4$time_new  +
    V2
  
  return(list(fit.S3S1 = fit.S3S1, fit.NHR = fit.NHR))
}
```

```{r}
fits = post.sample %>%
  lapply(function(x)
    fit.post.sample(x))

# then we average to get one fitted value for each time point.
S3S1.fit = fits %>%
  sapply(function(x)
    x$fit.S3S1) %>% 
  rowMeans()
S3S1.fit = S3S1.fit/100
NHR.fit = fits %>%
  sapply(function(x)
    x$fit.NHR) %>% 
  rowMeans()
```


```{r, echo=FALSE}
# the observed values
S3S1.obs = data.pt4$S3overS1
NHR.obs = data.pt4$NightHeartRate
library(ggplot2)
plot.df =
  cbind.data.frame(time = 1:116, S3S1.obs, S3S1.fit)
plot.pt4.S3S1 = plot.df %>%
  ggplot(aes(x = time)) +
  geom_line(aes(y = S3S1.obs), color = "#D55E00") +
  geom_line(aes(y = S3S1.fit), color = "#0072B2") +
  geom_vline(xintercept = n.train + 1,
             ylab = "",
             linewidth = 0.2) +
  xlab(label = "days") +
  ylab(label = "S3/S1") 
plot.pt4.S3S1
```




```{r}
n.train = n - n.hold
fore =c(rep(NA, n.train),
  tail(S3S1.fit, n.hold))
obs = S3S1.obs
fore.low <- c(
  rep(NA, n.train),
  tail(S3S1.fit, n.hold) -
    2 * tail(S3S1.fit, n.hold)
)
fore.up <- c(
  rep(NA, n.train),
  tail(S3S1.fit, n.hold) +
    2 * tail(S3S1.fit, n.hold)
)

plot.df <-
  cbind.data.frame(time = 1:n, obs, fore, fore.low, fore.up)

plot.S3S1.biv = plot.df %>%
  ggplot(aes(x = time)) +
  geom_line(aes(y = obs), color = "#D55E00") +
  geom_line(aes(y = fore), color = "#0072B2") +
  geom_ribbon(aes(ymin = fore.low, ymax = fore.up), alpha =
                .2) +
  geom_vline(xintercept = n.train + 1,
             ylab = "",
             size = 0.2) +
  xlab(label = "t") +
  ylab(label = expression(y[t])) +
  ggtitle("")
plot.S3S1.biv
```

```{r,echo=FALSE}
plot.df =
  cbind.data.frame(time = 1:116, NHR.obs, NHR.fit)
plot.pt4.NHR = plot.df %>%
  ggplot(aes(x = time)) +
  geom_line(aes(y = NHR.obs), color = "#D55E00") +
  geom_line(aes(y = NHR.fit), color = "#0072B2") +
  geom_vline(xintercept = n.train + 1,
             ylab = "",
             linewidth = 0.2) +
  xlab(label = "t") +
  ylab(label = expression(y[t])) +
  ggtitle("")
  plot.pt4.NHR
```

```{r}
fore =c(rep(NA, n.train),
  tail(NHR.fit, n.hold))
obs = NHR.obs
fore.low <- c(
  rep(NA, n.train),
  tail(NHR.fit, n.hold) -
     tail(NHR.fit, n.hold)
)
fore.up <- c(
  rep(NA, n.train),
  tail(NHR.fit, n.hold) +
    tail(NHR.fit, n.hold)
)

plot.df <-
  cbind.data.frame(time = 1:n, obs, fore, fore.low, fore.up)

plot.NHR.biv = plot.df %>%
  ggplot(aes(x = time)) +
  geom_line(aes(y = obs), color = "#D55E00") +
  geom_line(aes(y = fore), color = "#0072B2") +
  geom_ribbon(aes(ymin = fore.low, ymax = fore.up), alpha =
                .2) +
  geom_vline(xintercept = n.train + 1,
             ylab = "",
             size = 0.2) +
  xlab(label = "t") +
  ylab(label = expression(y[t])) +
  ggtitle("")
plot.NHR.biv
```
In this plot, we see the sensor for NHR. With blue is the fitted values and with red 
is the observed ones. Here, we see a better fit of the model as it follows nicely the 
observed ones. In the test set (last five measurements after the vertical black line)
we see a downward trend in the observed values which is also predicted from the model. 
This is an indication that the model for this sensor does a good job!

Lets calculate the MAE and MAPE for this model.
For S3/S1
```{r}
predicted.S3S1 = S3S1.fit[112:116]
test.S3S1 = test.df$S3overS1

mae = mae(test.S3S1, predicted.S3S1)
mae
```

```{r}
mape = mape(test.S3S1, predicted.S3S1)
mape
```

For NHR:
```{r}
predicted.NHR = NHR.fit[112:116]
test.NHR = test.df$NightHeartRate

mae = mae(test.NHR, predicted.NHR)
mae
```

```{r}
mape = mape(test.NHR, predicted.NHR)
mape
```


# FIT AR(2)


```{r}
formula.mvn.pt4.ar2 = Y ~ -1 + alpha.S3S1 + alpha.NHR + time.S3S1 + time.NHR + 
  f(t.index, model="iid2d", n=N) + 
  f(b0.S3S1, model = "ar", order = 2) +
  f(b0.NHR, model = "ar", order = 2)

inla.mvn.pt4.ar2 = inla(formula = formula.mvn.pt4.ar2,
                    family = c("gaussian"),
                    data = pt4.mvn.data,
                    control.inla = list(h = 1e-5, tolerance = 1e-3),
                    control.compute = list(dic = TRUE,waic = TRUE, cpo = TRUE,config = TRUE),
                    control.predictor = list(compute = TRUE),
                    control.family =list(hyper = list(prec = list(initial = 10,fixed = TRUE))))
result.ar2 = inla.rerun(inla.mvn.pt4.ar2)
summary(result.ar2)
```
```{r}
bri.hyperpar.summary(result.ar2)
```
```{r}
sigma.dist = inla.tmarginal(
  fun = function(x)
    exp(-x/2),
  marginal = result.ar2$marginals.hyperpar$`Precision for b0.NHR`
)
sigma.zm = inla.zmarginal(sigma.dist)

```
```{r}
# recover phis using the inverse transformation
pacf = result.ar2$summary.hyperpar$mean[5:6]
phi = inla.ar.pacf2phi(pacf)
print(phi)
```
```{r}
# recover phis using the inverse transformation
pacf = result.ar2$summary.hyperpar$mean[8:9]
phi = inla.ar.pacf2phi(pacf)
print(phi)
phi
```
```{r}
model.selection.criteria(result.ar2)
```
```{r}
#CPO and PIT
cpoi = result.ar2$cpo$cpo[!is.na(result.ar2$cpo$cpo)]
cpo = -sum(log(cpoi))
piti = result.ar2$cpo$pit[!is.na(result.ar2$cpo$pit)]
pit=ks.test(piti,"punif",0,1)$statistic

print(paste0("CPO: ", round(cpo,4)))
print(paste0("PIT: ", round(pit,5)))
```

```{r}
post.sample = inla.posterior.sample(1000, result.ar2)
```

Here is a function to calculate the fitted values
```{r}
fit.post.sample = function(x) {
  # firstly, we construct matrix V: V is a 2x2 covariance matrix 
  sigma2.v1 = 1 / x$hyperpar[1] # marginal variance for u_1
  sigma2.v2 = 1 / x$hyperpar[2] # marginal variance for u_2
  rho.v1v2 = x$hyperpar[3] # correlation
  cov.v1v2 = rho.v1v2 * sqrt(sigma2.v1) * sqrt(sigma2.v2) # covariance
  sigma.v =
    matrix(c(sigma2.v1, cov.v1v2, cov.v1v2, sigma2.v2), nrow = 2) # the covariance matrix
  
  # now we sample from the Multivariate Normal with mean 0 and covariance matrix the V
  V = rmvnorm(n, mean = c(0, 0), sigma = sigma.v)
  V1 = V[, 1] # first column for sensor 1 (S3/S1)
  V2 = V[, 2] # second column for sensor 2 (NHR)
  
  # using the M posterior samples, we simulate fitted responses for the 2 time series
  # For S3/S1
  fit.S3S1 = x$latent[grep("alpha.S3S1", rownames(x$latent))] +
    x$latent[grep("b0.S3S1", rownames(x$latent))] +
    x$latent[grep("time.S3S1", rownames(x$latent))] * 
    data.pt4$time_new +
    V1
  # for NHR
  fit.NHR = x$latent[grep("alpha.NHR", rownames(x$latent))] +
    x$latent[grep("b0.NHR", rownames(x$latent))] +
    x$latent[grep("time.NHR", rownames(x$latent))] * 
    data.pt4$time_new  +
    V2
  
  return(list(fit.S3S1 = fit.S3S1, fit.NHR = fit.NHR))
}
```

```{r}
fits = post.sample %>%
  lapply(function(x)
    fit.post.sample(x))

# then we average to get one fitted value for each time point.
S3S1.fit = fits %>%
  sapply(function(x)
    x$fit.S3S1) %>% 
  rowMeans()
S3S1.fit = S3S1.fit/100
NHR.fit = fits %>%
  sapply(function(x)
    x$fit.NHR) %>% 
  rowMeans()
```

```{r, echo=FALSE}
# the observed values
S3S1.obs = data.pt4$S3overS1
NHR.obs = data.pt4$NightHeartRate
plot.df =
  cbind.data.frame(time = 1:116, S3S1.obs, S3S1.fit)
plot.pt4.S3S1.ar2 = plot.df %>%
  ggplot(aes(x = time)) +
  geom_line(aes(y = S3S1.obs), color = "#D55E00") +
  geom_line(aes(y = S3S1.fit), color = "#0072B2") +
  geom_vline(xintercept = n.train + 1,
             ylab = "",
             linewidth = 0.2) +
  xlab(label = "days") +
  ylab(label = "S3/S1") 
plot.pt4.S3S1.ar2
```
```{r,echo=FALSE}
plot.df =
  cbind.data.frame(time = 1:116, NHR.obs, NHR.fit)
plot.pt4.NHR = plot.df %>%
  ggplot(aes(x = time)) +
  geom_line(aes(y = NHR.obs), color = "#D55E00") +
  geom_line(aes(y = NHR.fit), color = "#0072B2") +
  geom_vline(xintercept = n.train + 1,
             ylab = "",
             linewidth = 0.2) +
  xlab(label = "t") +
  ylab(label = expression(y[t])) +
  ggtitle("")
  plot.pt4.NHR
```

```{r}
predicted.S3S1 = S3S1.fit[112:116]
test.S3S1 = test.df$S3overS1
mae = mae(test.S3S1, predicted.S3S1)
mae
```

```{r}
mape = mape(test.S3S1, predicted.S3S1)
mape
```
For NHR:
```{r}
predicted.NHR = NHR.fit[112:116]
test.NHR = test.df$NightHeartRate
mae = mae(test.NHR, predicted.NHR)
mae
```

```{r}
mape = mape(test.NHR, predicted.NHR)
mape
```

# fit for all patients
```{r, echo=FALSE}
ids = data$Participant.Id
df = data.frame()
state.S3S1 = list()
state.NHR = list()
df.gof = data.frame()
for (i in unique(ids)){
  data.i = data[data$Participant.Id == i, ]
  data.i = data.i[data.i$time_new < 0,] # only the negative values
  
  n = nrow(data.i) # number of observations 
  N = 2*n
  n.hold = 5
  n.train = n - n.hold
  train.df = data.i[1:n.train, ]
  test.df = tail(data.i, n.hold)

  # S3/S1 ratio
  S3S1.pti = train.df$S3overS1*100
  # Night Heart Rate
  NHR.pti = train.df$NightHeartRate
  
  # create the Y vector
  Y = c(c(S3S1.pti, rep(NA, n.hold)), c(NHR.pti, rep(NA, n.hold)))
  
  # intercepts
  alpha.S3S1 = c(rep(1,n), rep(NA,n))
  alpha.NHR = c(rep(NA,n), rep(1,n))
  
  # trends
  time.S3S1 <- c(data.i$time_new, rep(NA, n))
  time.NHR <- c(rep(NA, n), data.i$time_new)
  
  # AR(1) processes
  b0.S3S1 = c(1:n, rep(NA, n))
  b0.NHR = c(rep(NA, n), 1:n)
  
  # for covariance matrix
  t.index = 1:N
  pti.mvn.data = cbind.data.frame(Y,
                       alpha.S3S1, alpha.NHR,
                       b0.S3S1, b0.NHR,
                       time.S3S1, time.NHR,
                       t.index)
  
  formula.mvn.pti = Y ~ -1 + alpha.S3S1 + alpha.NHR + time.S3S1 + time.NHR + 
  f(t.index, model="iid2d", n=N) + 
  f(b0.S3S1, model = "ar1") +
  f(b0.NHR, model = "ar1")

  inla.mvn.pti = inla(formula = formula.mvn.pti,
                      family = c("gaussian"),
                      data = pt4.mvn.data,
                      control.inla = list(h = 1e-5, tolerance = 1e-3),
                      control.compute = list(dic = TRUE,waic = TRUE, cpo = TRUE,config = TRUE),
                      control.predictor = list(compute = TRUE),
                      control.family =list(hyper = list(prec = list(initial = 10,fixed = TRUE))))
  result = inla.rerun(inla.mvn.pti)


  gof.dic = result$dic$dic
  gof.waic = result$waic$waic
  gof.ml = result$mlik
  
  output.int.S3S1 = ar2$summary.fixed$mean[1]
  output.int.NHR = ar2$summary.fixed$mean[2]
  
  output.slope.S3S1 = ar2$summary.fixed$mean[3]
  output.slope.NHR = ar2$summary.fixed$mean[4]
  
  output.ar1.S3S1 = ar2$summary.hyperpar$mean[5]
  output.ar1.NHR = ar2$summary.hyperpar$mean[7]
  
  state.S3S1 = append(state.S3S1,list(result$internal.marginals.hyperpar$`Log precision for b0.S3S1`))
  state.NHR = append(state.NHR,list(result$internal.marginals.hyperpar$`Log precision for b0.NHR`))

  
  output = c(output.int.S3S1, output.int.NHR, output.slope.S3S1, output.slope.NHR, output.ar1.S3S1, output.ar1.NHR)
  output.gof = c(gof.dic, gof.waic, gof.ml)

  df = rbind(df, output)
  df.gof = rbind(df.gof, output.gof)
}
colnames(df) = c("Level_S3S1", "Level_NHR", "Trend_S3S1", "Trend_NHR", "AR1_S3S1", "AR1_NHR")
df$ID = unique(ids)

colnames(df.gof) = c("DIC", "WAIC", "ML1", "ML2")
df.gof$ID = unique(ids)
```
